how to \texttt{git bisect} when your test suite is having a bad day

### 2020-03-30 bisect/paper/paper.tex
finish this dang bisect project

\keywords pdflatex, cdflatex
% bracket, groove, in, jumps, the

\newcommand\confidents{\ensuremath{\mathcal{Z}}\xspace}
\newcommand\pdf{\ensuremath{\mathsf{pdf}}\xspace}
\newcommand\cdf{\ensuremath{\mathsf{cdf}}\xspace}

@@

If you're feeling adventurous,
implement a strategy and throw it in this simulator I implemented to see how it stacks up:
\url{https://github.com/bblum/sigbovik/blob/master/bisect}.
You just gotta implement {\tt trait BisectStrategy},
and it even does all the hard work of applying Bayes's rule
%computing the
%distribution of
%priors with Bayes's rule
for you and letting you see the PDF and everything.
Check it out.

@@

\section{pdfs, but not the portable document format kind, and our friend rev. bayes}

@@

Now let's say the test passes at $j=9$, then at $j=10$.
\cref{fig:example}(b) shows the updated PDFs/CDFs:
for $p=1$, this pinpoints the bug at $j=11$, and the search is over.
But for $p=0.5$, there's still 2/3 odds we'd be wrong!
In fact, from here it takes 18 further probes at $j=10$ until we are at least five 9's confident that $c_{11}$ is the culprit.%
\footnote{See {\tt fn test\_figure\_1()}!}
Bayes's rule gonna get ya.

@@

Dropbox has a CI service called Athena \cite{athena}
which automatically searches for flaky tests. %, when otherwise no bug exists.
In this case the goal is to keep the build green,
but if you consider the flaky test itself to be the bug, it's the same problem.%
\footnote{Incidentally, the symmetric case
-- where a bug repros with $p=1$, but the test also flakes with some $q<1$ --
is also the same problem.}
Athena's ``deflakes'' the test at each commit by running 10 times,
treating the combined result as ``basically as good as $p=1$'',
and then runs an otherwise classical binary search.
%\footnote{Upcoming, I will call this strategy \textsf{Human-Mistrustful}$(10)$.}
In this setting, $p$ is not known in advance, so using Bayes's rule is off the table.
But I will show that even without access to the PDF, a better strategy exists. % at least by my definition

@@

Another way to frame it is that the $c_{(a+b)/2}$ bisect point
% maximizes the expected probability mass that moves from one side to the other.
cuts the probability mass, rather than the range, in half,
% How about in Nondeterminism World,
% we ignore the range length, and instead bisect at the midpoint of probability mass,
i.e., $\mathsf{max}_j (i)$ s.t. $\cdf(j) \le 0.5)$.
This approach
fits the ``binary search'' spirit,
and will terminate in Nondeterminism World,
i.e.,
if $b$ is the solution, it converges to repeatedly probing $b-1$,
so it can pass any fixed $\confidents$ threshold.
%(which is more than you can say for the range-based $c_{(0+b)/2}$).
%It is intuitively obvious that this will make progress faster than bisecting at $c_{(b+0)/2}$,
%But is it fast?
% However, it doesn't maximize either the minimum or the expected amount of mass that moves:
% when $p<1$, it's intuitively apparent that this CDF threshold should actually be less than $0.5$.%
%\footnote{I did some pen-and-paper algebra to find a closed form for this,
%but I actually trust the simulator more than myself at this point,
%so I stopped spending brains on this part of the problem. I kind of regret that.}
%
%Is maximizing the expected moved mass as a function of $p$ the fastest we can go,
%and if so, why?
But is $0.5$ still the best bisect point even when $p<1$?
I wondered if this could be expressed as the amount of probability mass moved from one side to the other,
i.e., $\sum_i \mathsf{abs}(\pdf(i) - \pdf'(i))$.
In the case where the bug repros this is:
\[
	p\cdf(j) \times (1-\cdf(j))
\]
and in the case where the test passes:
\[
	(1-p\cdf(j)) \times \cdf(j) \times (1-\renorm{i\le j})
\]
which, surprisingly, simplifies to exactly the same thing as the bug repros case.
The maximum occurs where $\partial/\partial \cdf(j)$ is 0,
which turns out to be at $\cdf(j)=0.5$ after all, and independent of $p$.%
\footnote{I also worked out the {\it minimum} moved mass between the pass and fail case,
which is almost always the pass case.
It comes out to
%$\frac{p\cdf(j)(1-\cdf(j))}{1-p\cdf()}$
$\frac{\cdf(j)(1-\cdf(j))}{1/p-\cdf(j)}$,
which experiences its maximum at
$\frac{1-\sqrt{1-p}}{p}$ (thanks wolframalpha).
But this is worst-case thinking, which doesn't seem appropriate.
Let's have some optimism!}
But it's not clear that the amount of mass moved
necessarily corresponds to reaching $\confidents$ the fastest.
%It just sounds kinda nice?
%What is the best threshold, perhaps as a function of $p$, and why?
I show my work in \cref{fig:let-me-shown-that-my-work},
because that's what they taught me to do in high school.

@@

There is a ray of hope, however:
thanks to the nondecreasing-PDF invariant I mentioned on the previous page,
the expected entropy has a unique local minimum.%
\footnote{See {\tt fn test\_expected\_entropy\_has\_unique\_local\_minimum()}.}
Thus we can binary search for the global minimum, making this at least $O(n\mathsf{log}n)$
instead of $O(n^2)$.

@@

The most straightforward thing for a human to do is to assume $p=1$ until they encounter evidence to the contrary.
They'll conduct a normal binary search, and when they think they've found the bug at $c_b$,
they'll just keep probing $c_{b-1}$ until enough probability mass (invisible to them)
moves to $b$ and the simulator says they can stop.
If this instead repros the bug at $c_{b-1}$ after all, the human gets confused!
They'll assume the bug might have been present as early as $c_0$,
forgets their lower bound, and begins binary searching anew in the range $[0,b-1]$.
It's easy to see this makes progress.
Let's call this the ``confused, forgetful human'' strategy.

Another confused human variant is for them to remember their previous lower bounds.
After all, just because their best lower bound $b-1$ was contradicted
doesn't say anything about some earlier bound $a<b-1$ they might have observed.
So this human will maintain a stack of lower bounds, and when $b-1$ is contradicted,
they'll pop off the stack and resume searching in $[a,b-1]$ instead.
Let's call this the ``confused human who remembers''.
%
Incidentally, while implementing this strategy, I accidentally wrote the comment,
\begin{center}
	\begin{tabular}{l}
		{\tt // the human, who now has perfect memory,} \\
		{\tt // walks backwards in time}
	\end{tabular}
\end{center}
and giggled a lot to myself as I imagined {\birbaslo sigbovik}~fanfiction.

Finally, let's imagine the human knows maybe something is up with this $p$ stuff in advance. %before they start searching.
Wishing to carry on classically as though $p$ were 1,
they'll try to emulate $p$ being higher by retrying any passing test ``just to be sure'', before moving on.
If they retry the test $r$ times,
the probability it fails given the bug is present is then $1-(1-p)^{1+r}$.
%In general, for $r$ retries, that's $1-(1-p)^{r+1}$.
But it comes at the cost of $r$ more steps for each new lower bound!
As before, when this human thinks they're done, they'll repeatedly probe $c_{b-1}$ until contradicted
(and we'll make them forgetful, to keep things simple).
Let's call this the ``suspicious human of $r$ retries''.

@@

To save you some page-turning, here's the source code link again:
\url{https://github.com/bblum/sigbovik/blob/master/bisect}.
A {\tt BisectStrategy}
implements a function over $p$, the PDF, and/or its own internal state
to choose the next bisect point.
The {\tt SimulationState} is initialized with fixed $n$ and $p$,
and repeatedly invokes a given {\tt BisectStrategy},
hitting the PDF with Bayes's rule accordingly,
until $\confidents > 0.99999$.
I provide implementations for all the strategies from the previous section in {\tt src/strategies/}.
It's written in Rust so it's self-documenting.

I learned a lot about floating point.
I don't mean like NaNs and mantissa bits,
I mean that when your project is fundamentally a long multiplication chain of increasingly high rationals,
your precious $\cdf(n)$ inevitably drifts farther from 1 than you can bound with any arbitrary $\epsilon$, and you start drowning in imprecision \cite{drowningpoint}.
I suppose I could have used the {\tt BigRational} library,
but I chose Rust over Haskell for this so I could {\it avoid} runaway memory consumption...
so I resigned myself to explicitly renormalizing my PDFs by $1/\cdf(n)$ each time I updated them.
After that, I was able to write some assertions
to bound the imprecision drift within $k(\text{\tt std::f64::EPSILON})$ \cite{epsilon}.
But yeah, I definitely spent some sanity points writing a function named {\tt fn assert\_kinda\_equals\_one(\&self)}.

@@

Of course, each trial takes exponentially
%\footnote{Yes, I am using this word in the literal sense.}
longer and longer as $p \rightarrow 0$, so I didn't bother testing past 0.1.
I also didn't bother measuring execution time,
as prior work has shown that one's personal laptop is a fraught environment for stable performance evaluation \cite{htm},
but it was still very obvious that {\sf entropy} was far, far slower per step than all other strategies.
%I guess I'd say it took minutes to run {\sf cdfbisect} up to $p=0.01$, and hours/overnight for the same for {\sf entropy}.

@@

\cref{fig:resulce} was getting a little crowded so I've plotted all the human-friendly strategies on \cref{fig:human}.
It's pretty self-explanatory imo.

@@

I had the simulator refuse to count a wrong result, %towards the 65536 trial count,
restarting the search until it correctly identified the buggy commit 65536 times,
and counting the number of restarts.
With $\confidents > 0.99999$ this was not very interesting:
the strategies collectively made 0 errors 81\% of the time, and were wrong more than twice only 0.6\%.
But $\confidents > 0.5$ was a different story.
As \cref{fig:wrongs} shows, most strategies
-- even the comically slow {\sf rand-uniform} --
converged to being wrong about half the time, i.e., 65536 wrongs per 65536 rights.
{\sf athena} was most resilient, at least for higher $p$.

>>>>

\keywords pdflatex, cdflatex, epsilon
% bracket, groove, in, jumps, the

\newcommand\confidents{\ensuremath{z}\xspace}
\newcommand\pdf{\ensuremath{\mathsf{pdf}}\xspace}
\newcommand\cdf{\ensuremath{\mathsf{cdf}}\xspace}

@@

If you're feeling adventurous,
implement a strategy and throw it in this simulator I made to see how it fares:
\url{https://github.com/bblum/sigbovik/blob/master/bisect}.
You just gotta implement {\tt trait BisectStrategy},
and it even does all the hard work of applying Bayes's rule
%computing the
%distribution of
%priors with Bayes's rule
for you and letting you see the PDF and everything.
Check it out.

@@

\section{pdfs, but not the portable document format kind, and our friend rev. bayes}
\label{sec:pdfs}

@@

Now let's say the test passes at $j=9$, then at $j=10$.
\cref{fig:example}(b) shows the updated PDFs/CDFs:
for $p=1$, this pinpoints the bug at $j=11$, and the search is over.
But for $p=0.5$, there's still 2/3 odds we'd be wrong!
In fact, from here it takes 18 further probes at $j=10$ until we are at least five 9's confident that $c_{11}$ is the culprit.%
\footnote{See {\tt fn test\_figure\_1()}.}
Bayes's rule gonna get ya.

@@

Dropbox has a CI service called Athena \cite{athena}
which automatically searches for flaky tests. %, when otherwise no bug exists.
In this case the goal is to keep the build green,
but if you consider the flaky test itself to be the bug, it's the same problem.%
\footnote{Incidentally, the symmetric case
-- where a bug repros with $p=1$, but the test also flakes with some $q<1$ --
is also the same problem.}
Athena ``deflakes'' the test at each commit by running it 10 times,
treating the combined result as ``basically as good as $p=1$'',
and then runs an otherwise classical binary search.
%\footnote{Upcoming, I will call this strategy \textsf{Human-Mistrustful}$(10)$.}
In this setting, $p$ is not known in advance, so using Bayes's rule is off the table.
But I will show that even without access to the PDF, a better strategy exists. % at least by my definition

@@

Another way to frame it is that the $c_{(a+b)/2}$ bisect point
% maximizes the expected probability mass that moves from one side to the other.
cuts the probability mass, rather than the range, in half,
% How about in Nondeterminism World,
% we ignore the range length, and instead bisect at the midpoint of probability mass,
i.e., $\mathsf{max}_j (j$; $\cdf(j) \le 0.5)$.
This approach
fits the ``binary search'' spirit,
and will also terminate in Nondeterminism World:
%i.e.,
if $b$ is the solution, it converges to repeatedly probing $b-1$,
so it can pass any fixed $\confidents$ threshold.
%(which is more than you can say for the range-based $c_{(0+b)/2}$).
%It is intuitively obvious that this will make progress faster than bisecting at $c_{(b+0)/2}$,
%But is it fast?
% However, it doesn't maximize either the minimum or the expected amount of mass that moves:
% when $p<1$, it's intuitively apparent that this CDF threshold should actually be less than $0.5$.%
%\footnote{I did some pen-and-paper algebra to find a closed form for this,
%but I actually trust the simulator more than myself at this point,
%so I stopped spending brains on this part of the problem. I kind of regret that.}
%
%Is maximizing the expected moved mass as a function of $p$ the fastest we can go,
%and if so, why?
But is $0.5$ still the best bisect point even when $p<1$?
I wondered if this could be expressed as the amount of probability mass moved from one side to the other,
i.e., $\sum_i \mathsf{abs}(\pdf(i) - \pdf'(i))$.
In the case where the bug repros this is:
\[
	p\cdf(j) \times (1-\cdf(j))
\]
and in the case where the test passes:
\[
	(1-p\cdf(j)) \times \cdf(j) \times (1-\renorm{i\le j})
\]
which, surprisingly, simplifies to exactly the same thing as the bug repros case.
The maximum occurs where $\partial/\partial \cdf(j)$ is 0,
which turns out to be at $\cdf(j)=0.5$ after all, and independent of $p$.%
\footnote{I also worked out the {\it minimum} moved mass between the pass and fail case,
which is almost always the pass case.
It comes out to
%$\frac{p\cdf(j)(1-\cdf(j))}{1-p\cdf()}$
$\frac{\cdf(j)(1-\cdf(j))}{1/p-\cdf(j)}$,
which experiences its maximum at
$\frac{1-\sqrt{1-p}}{p}$ (thanks wolframalpha).
But this is worst-case thinking, which doesn't seem appropriate.
Let's have some optimism!}
But it's not clear that the amount of mass moved
necessarily corresponds to reaching $\confidents$ the fastest.
%It just sounds kinda nice?
%What is the best threshold, perhaps as a function of $p$, and why?
I show my work in \cref{fig:let-me-shown-that-my-work},
because that's what they taught me to do in high school algebra class.

@@

There is a ray of hope, however:
thanks to the nondecreasing-PDF invariant I mentioned in \cref{sec:pdfs}, %on the previous page,
the expected entropy has a unique local minimum.%
\footnote{See {\tt fn test\_expected\_entropy\_has\_unique\_local\_minimum()}.}
Thus we can binary search for the global minimum, making this at worst $O(n\mathsf{log}n)$
instead of $O(n^2)$.

@@

The most straightforward thing for a human to do is to assume $p=1$ until they encounter evidence to the contrary.
They'll conduct a normal binary search, and when they think they've found the bug at $c_b$,
they'll just keep probing $c_{b-1}$ until enough probability mass (invisible to them)
moves to $b$ and the simulator says they can stop.
If this instead repros the bug at $c_{b-1}$ after all, the human gets confused!
They'll assume the bug might have been present as early as $c_0$,
forgets their lower bound, and begins binary searching anew in the range $[0,b)$.
It's easy to see this makes progress.
Let's call this the ``confused, forgetful human'' strategy.

Another confused human variant is for them to remember their previous lower bounds.
After all, just because their best lower bound $b-1$ was contradicted
doesn't say anything about some earlier bound $a<b-1$ they might have observed.
So this human will maintain a stack of lower bounds, and when $b-1$ is contradicted,
they'll pop off the stack and resume searching in $[a,b)$ instead.
Let's call this the ``confused human who remembers''.
%
Incidentally, while implementing this strategy, I accidentally wrote the comment,
\begin{center}
	\begin{tabular}{l}
		{\tt // the human, who now has perfect memory,} \\
		{\tt // walks backwards in time}
	\end{tabular}
\end{center}
and giggled a lot to myself as I imagined {\birbaslo sigbovik}~fanfiction.

Finally, let's imagine the human knows in advance that maybe something is up with this $p$ stuff.
%in advance. %before they start searching.
Wishing to carry on classically as though $p$ were 1,
they'll try to emulate $p$ being higher by retrying any passing test ``just to be sure'', before moving on.
If they retry the test $r$ times,
the probability it fails given the bug is present is then $1-(1-p)^{1+r}$.
%In general, for $r$ retries, that's $1-(1-p)^{r+1}$.
But it comes at the cost of $r$ more steps for each new lower bound!
As before, when this human thinks they're done, they'll repeatedly probe $c_{b-1}$ until contradicted
(and we'll make them forgetful, to keep things simple).
Let's call this the ``suspicious human of $r$ retries''.

@@

To save you some page-turning, here's the source code link again:
\url{https://github.com/bblum/sigbovik/blob/master/bisect}.
A {\tt BisectStrategy}
implements a function over $p$, the PDF, and/or its own internal state
to choose the next bisect point.
The {\tt SimulationState} is initialized with fixed $n$ and $p$,
and repeatedly invokes a given {\tt BisectStrategy},
hitting the PDF with Bayes's rule accordingly,
until $\confidents > 0.99999$.
I provide implementations for all \cref{sec:strats}'s strategies in {\tt src/strategies/}.
It's written in Rust so it's self-documenting.

I learned a lot about floating point.
I don't mean like NaNs and mantissa bits,
I mean that when your project is fundamentally a long multiplication chain of increasingly high rationals,
your precious $\cdf(n)$ inevitably drifts farther from 1 than you can bound with any arbitrary $\epsilon$, and you start drowning in imprecision \cite{drowningpoint}.
I suppose I could have used the {\tt BigRational} library,
but I chose Rust over Haskell for this so I could {\it avoid} runaway memory consumption...
so, I resigned myself to explicitly renormalizing my PDFs by $1/\cdf(n)$ each time I updated them.
After that, I was able to write some assertions
to bound the imprecision drift within $k(\text{\tt std::f64::EPSILON})$ \cite{epsilon}.
But yeah, I definitely spent some sanity points writing a function named {\tt fn assert\_kinda\_equals\_one(\&self)}.

@@

Of course, each trial takes exponentially
%\footnote{Yes, I am using this word in the literal sense.}
longer and longer as $p \rightarrow 0$, so I didn't bother testing past 0.1.
I also didn't bother measuring execution time,
as prior work has shown that one's personal laptop is a fraught environment for stable performance evaluation \cite{htm},
but it was still very obvious that {\sf entropy} was far, far slower per step than all other strategies.
%I guess I'd say it took minutes to run {\sf cdfbisect} up to $p=0.01$, and hours/overnight for the same for {\sf entropy}.
It was even slower than {\sf rand-uniform}!

@@

\cref{fig:resulce} was getting a little crowded so I've plotted all the human-friendly strategies separately on \cref{fig:human}.
It's pretty self-explanatory imo.

@@

I had the simulator refuse to count a wrong result, %towards the 65536 trial count,
restarting the search until it correctly identified the buggy commit 65536 times,
and counting the number of restarts.
With $\confidents > 0.99999$ this was not very interesting:
the strategies collectively made 0 errors 81\% of the time, and were wrong more than twice only 0.6\%.
But $\confidents > 0.5$ is a different story.
As \cref{fig:wrongs} shows, most strategies
-- even the comically slow {\sf rand-uniform} --
converged to being wrong about half the time, i.e., 65536 wrongs per 65536 rights.
For higher $p$,
{\sf athena} was most resilient.
